Characterization of gaps in directly bonded Si compound optics

### 2014-02-13 AO_bondedSi.tex
add prose on IR imaging

The University of Texas diffractive optics group produces diffraction gratings on silicon using microlithographic techniques \cite{1998SPIE.3354..201J, 2010SPIE.7739E.146W}.  These gratings are used either in transmission (i.e. as grisms \cite{2008SPIE.7014E..77D, 2010SPIE.7739E.123G}), or in reflection (i.e. as immersion gratings, \cite{2007ApOpt..46.3400M, 2010SPIE.7739E.146W, 2012SPIE.8450E..2SG}.  These gratings operate throughout the infrared, from 1.2 to 38 $\mu$m, and find application in ground-based (IGRINS \cite{2010SPIE.7735E..54Y, 2012SPIE.8450E..2SG}), airborne (SOFIA FORCAST \cite{2008SPIE.7014E..77D}), and space (JWST NIRCAM \cite{2005SPIE.5904...21B, 2010SPIE.7739E.123G}) applications.  To date, our group has produced the gratings on large (up to 100 mm diameter and 30 mm thick) monolithic pieces of monocrystalline silicon and we shape the substrates by sawing, grinding, and polishing to form the final devices \cite{2010SPIE.7739E.146W}.  The next generation of astronomy and earth science spectrographs, however, will need optical devices larger and more precise than what can be made with existing technology.  The proposed Giant Magellan Telescope\cite{2012SPIE.8444E..1HJ} Near Infrared Spectrograph\cite{2006SPIE.6269E.143J, 2010SPIE.7735E..87L} (GMTNIRS) will need immersion gratings with thicknesses up to 60 mm and lengths up to 200 mm.  Substrates of the required size cannot fit into conventional processing equipment like reactive ion etchers and electron-beam lithography tools.  Large, thick Si substrates are also prohibitively expensive to experiment on.  Silicon direct bonding\cite{1986JAP....60.2987S, 2012SPIE.8450E..2TV} offers advantages by reducing the thickness and mass needed for the development of a thin (0.5$-$10 mm) Silicon diffractive optic that can later be stuck to a fine polished Si substrate, from which an immersion grating prism can be made.  Other groups producing SI diffractive optics are also experimenting with this approach (see \cite{2012SPIE.8450E..2TV} and \url{www.lumella.com}).

While it is possible that proper precautions to eliminate all micron-sized particles from both SI surfaces prior to bonding, the power-law nature of the particle size distribution in clean-room environments (\cite{doi:10.1080/02786828608959094}) means that it is significantly more difficult to assure that no particles of the 10$-$60 nm size scale are present.

There are two major concerns with Si direct bonding: the performance and robustness.  Specifically, what happens if the bonding process is imperfect and there are small gaps between the Si parts?  Does a gap degrade the performance of a final device, either through optical aberrations or transmission loss?  How mechanically robust is the final unit?  What is the strength of the bond?  Can the interface be broken by shearing, torquing, or pulling?  In practice, the figure of merit for robustness is whether the part can survive a shake test simulating the forcing experienced during a rocket launch or transport, or during cryogenic cycling, if the part will be used at low temperature.  Optically, the relevant parameters are the loss caused by the gap and the effective distortion of the transmitted wavefront.

We describe a technique for directly measuring gaps at silicon-silicon interfaces.  Other techniques for measuring and mapping gaps have already been described elsewhere\cite{1992JEMat..21..669M}, including IR interference, X-ray topography, and ultrasound microscopy.  The standard infrared interference technique is to examine the bond in reflection for the presence of Newton rings.  The presence and number of such rings can yield information about both the lateral and axial extend of any gap, as long as the gap along the direction of propagation is of the order of the vacuum wavelength or larger.  X-ray topography \cite{1992JEMat..21..669M, 1994JaJAP..33....6H} exploits the change in optical density along lines of sight with gaps to allow precise and high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but provide only coarse estimates of the axial extent.  Ultrasound microscopy exploits the reflections at the gap interface to produce reflections of sound waves and map the lateral gap dimensions but produces limited information about the axial dimension \cite{2000RScI...71.1869G}.

Here we describe an exploitation of the IR interference technique with new precision instrumentation in the near infrared wavelength range to measure gaps at the few nanometer size scale.  In this paper we show that gaps as small as 35 nm cause an unacceptable degradation in transmission performance.  Because the size scales probed are so small, they are comparable to the range over which molecular (van der Waals) forces are acting.  It is conceivable that the strength of the bond energy can be constrained from the \emph{in situ} measurement of the separation and that therefore the optical technique can provide not only a measure of the transmission but also an indication of the mechanical robustness of the bond.  By measuring both the spatial and axial extent of any gaps the new metrology technique informs revisions of the bonding strategy that can improve both the optical and mechanical performance.

Our new technique for measurement of the axial extent of gaps exploits wavelength-dependence of reflection losses at the gap and the availability of commercial scanning monochromator systems with both high precision and good amplitudes stability.  We first describe the theory behind the technique and predicted outcomes for various gap dimensions.  We then evaluate the technique on a variety of bonded samples.

\section{History of the IR imaging technique}

@@

See the table below for a list of papers on Si wafer bonding and a note on their contribution.

@@

\begin{center}
\begin{longtable}{ |c | p{8cm} | p{8cm} |}
%\caption{History of Si direct bonding and metrology \label{tab:litbonding}}

%    \begin{tabular}{ |c | p{8cm} | p{8cm} |}

@@

    \hline
%    \end{tabular}
\end{longtable}
\end{center}

@@

The measurement technique exploits the large Fresnel reflection \cite{2001opt4.book.....H} at the Si-Air interface.  The refractive index of silicon at room temperature ranges from 3.55$-$3.45 from $\lambda = $ 1150$-$2500 nm\cite{2006SPIE.6273E..77F}, one of the largest refractive indices of any conventional material.  The Fresnel reflection is accordingly about 30\% per surface.  Popular optical materials with refractive indices in the range 1.5$-$1.7 with Fresnel reflection 4$-$7\% per surface will suffer less than Si from the effects described here.

Consider the following two conceivable pitfalls in Si direct bonding.  First, particles at the tens of nanometer size scale are trapped between the Si pucks during bonding.  This potential problem is especially acute for silicon direct bonding compared to bonding of popular low-index optical materials.  A typical particle surface flux is shown in Figure 5 of Cooper 1986\cite{doi:10.1080/02786828608959094}.  The flux of particles is a step function of particle size.  For every micron sized particle causing Newton's rings, there will be as many as 100,000 times more ten-nanometer sized particles.

The second pitfall to consider is that the stiff non-flat Si surfaces will not conform upon contact leaving the complement of the aberrations as an air gap. Similarly, surface roughness will also limit the bonding strength  \cite{JJAP.37.4197, 2001JOptA...3...85G}.  The details of this aspect of the bonding process are outside the scope of this paper.

\section{Substrate preparation}
The substrates were silicon wafers, some had unknown pedigree.  The surface roughness was typically about 2 nm, as measured with a Wyko 2D noncontact microscope.  The substrate thicknesses ranged from 0.5 to 3.3 mm per substrate.  We classify the 3.3 mm thick substrates as ``pucks'' and the $\sim$1.0 mm thick substrates as ``wafers''.  The distinction between the two classes of substrate thickness is based on the anticipation that thin substrates will be more conformal than thick ones.  There is also evidence that bond front propagation is much slower in thick pucks \cite{2007ApOpt..46.6793H} than in wafers.  We measured the large scale surface flatness of the pucks.  The Fisba2 interferometer has a 2 inch diameter beam at $\lambda=$632.8 nm.  We found a peak to valley surface flatness of 2 waves over the central 2 inch diameter.

@@

\begin{table}[h!]
\caption{UTexas Si bonding experiments \label{tab:bondexper}}
\begin{center}
    \begin{tabular}{ c c c c c c c}
    \hline
    Name & Bonded/DSP & Diameter & Thickness & Anneal Temp & Date & Pattern notes \\ 
    - & - & mm & mm & $^\circ$C & MM/DD/YYY & - \\ 
        \hline
    VG01 & Bonded  & 100  &$1$ $\times$ 2 & 23  & 2/13/2013 &  \\
    VG02 & Bonded & 100 & $3.3$ $\times$ 2 &  23  & 2/13/2013 &  \\
    VG03 & Bonded & 100 & $3.3$ $\times$ 2 &  23 & 2/14/2013 & 4 $\mu$m hole \\    
    VG04 & Bonded & 100 & $1$ $\times$ 2 &  23 & 2/14/2013 & 15 nm gap \\        
    VG05 & DSP & 100 & 1&  23 & 2/15/2013 & cleaved  \\
    VG06 & DSP & 100 & 3.5 &  23 & 2/15/2013 & \\
    VG07 & - & - & - &  - & - & \\
    VG08 & - & - & - &  - & - & \\
    VG09 & Bonded & 75 & 1.5? & 23 & 6/21/2013 & mesh C, 49 $\pm$6 nm \\
    VG15 & Bonded & 100 & 1.5? &  23 & 6/21/2013 & mesh C, M, F, 14 $\pm$2 nm\\
    VG17 & Bonded & 75 & 6.0? &  23 & 6/21/2013 & mesh C, 64 $\pm$6 nm\\
    VS20 & Bonded & 100 & 1.5? &  23 & 6/21/2013 & mesh C, M, F, 95 $\pm$5 nm\\
    \hline
    \end{tabular}
\end{center}
\end{table}

\section{Transmission measurement theory}
The Si-air interfaces of the air gap make a Fabry-P\`erot etalon\cite{2007fuph.book.....S}.  The transmission through a Fabry-Perot depends on the wavelength of light, the reflectivity of the etalon sidewalls, and the size of the gap.  We assume transmission is at normal incidence, the index of air is 1.0, and optics are at room temperature.  We assume Si is nonabsorbing longward of $\lambda$ = 1150 nm.  We compute the reflectivity for an Si-air interface, which is a modest function of wavelength through the refractive index of Silicon, $n(\lambda, T)$ \cite{2006SPIE.6273E..77F}.  The reflectivity is about 30\% in the wavelength range 1100 to 5000 nm.  The coefficient of Finesse\cite{2007fuph.book.....S} encapsulates the Fabry-P\`erot etalon's dependence on the reflectivity.  We compute a coefficient of Finesse in the range of about 2.8$\pm$0.1 depending on the wavelength.

\subsection{Consideration of multiple incoherent reflections}
Multiple incoherent reflections occur within the Si substrate.  Treatment for multiple coherent reflections have been described in detail in the optics literature \cite{2007fuph.book.....S}.  The wave transfer matrix technique treats each dielectric interface as a matrix with elements relating the pre- and post- interface complex amplitudes in the left and right directions.  I adapted the wave transfer technique for incoherent interactions \cite{2002ApOpt..41.3978K}.  Specifically I constructed an incoherent wave transfer matrix whose elements relate the intensities (and not complex amplitudes) before and after an interface.  The details are worked out in Appendix \ref{sec:Append-IMRTMM}.  We computed the transmission for two scenarios- first a double side polished (DSP) Si puck with no air gap, and second a pair of bonded Si pucks with an air gap thickness $d$.  We normalize the transmission with a gap to the transmission with no gap to isolate the effect of the gap.  We call this normalized transmission the etalon transmission $T_{e}$.  It is derived in Equation \ref{eqn:Tetalon} in the Appendix:

\begin{eqnarray}
T_{e} = \frac{n^2+1}{2 n F \sin ^2(2\pi \frac{d}{\lambda})+n^2+1} \nonumber
\end{eqnarray}

@@

\section{Measurements}
The measurements were performed with an Agilent Cary 5000 UV-Vis-NIR spectrophotometer \url{cary5000web}.  The typical measurement parameters and tool performance are listed below.  More details about the tool are available at the vendor's website.  Briefly, the tool has a double monochromator.  The double monochromator feature reduces scattered light, with the vendor claiming 40 dB of linearity.  We independently checked the linearity in the following way.  We separately measured the transmission spectrum of 5 neutral density filters.  We computed the mean transmission of each filter, predicting the net transmission of a combination of filters up to all five filters.  We then measured the net spectrum and compared to prediction.  The measured mean transmission of the stacks were indistinguishable from the predicted mean transmission within the measurement uncertainty over 24 dB.

The Cary 5000 has four operation modes: single front, single back, double, and double reverse.  We experimented with single front and double modes.  We estimated the uncertainty and measurement repeatability of single and double modes in the following way.  We computed the RMS error and measurement repeatability in a single measurement by computing the standard deviation of a the beam with no sample in a series of spectra taken immediately after a 100\% baseline scan.  The predicted mean level should naturally be 100\%.  We found that the mean value in single mode drifted by about 0.2\%.  The standard deviation of the featureless spectrum was typically 0.02\%, with some portions of the spectrum demonstrating increased uncertainty, perhaps attributable to atmospheric absorption. In double beam mode the spectrograph performance was much better.  We put a double side polished Si wafer in the reference beam, and no sample in the target beam during the baseline scan.  In this way the transmission spectrum of a target Si wafer is automatically normalized.  The double beam mode effectively divides out variations in the mean transmission levels attributable to lamp drift at 0.2\% in single beam measurements.  The drift in mean transmission level is merely 0.02\% in double beam mode.  These values are accurate for measurements over a half hour timescale or less.  Over periods of hours the drift increases for both single and double beam mode, so a repeat baseline scan is sometimes necessary.  The need for repeat baselines is more acute for single mode.

@@

\begin{table}[h!]
\caption{Summary of Cary 5000 Measurement parameters}
\begin{center}
    \begin{tabular}{ c c }
    \hline
    Parameter (Units) & Value \\ 
        \hline
       Spectral sampling interval (nm) & 2.0 \\
       Spectral resolution (nm) & 5 \\
       Time per sample (s) & 0.1 \\
       Available measurement range (nm) & 175$-$3300 \\
       Typical measurement range (nm) & 1000$-$2500 \\
	Single mode RMS  (\%) & 0.02 \\
	Double mode RMS  (\%) & 0.002 \\
	Single mode mean drift (\%) & 0.20 \\
	Double mode mean drift (\%) & 0.02 \\
	Mean 0\% baseline correction (\%) & 0.03 \\
	Beam size (mm $\times$ mm) & 2 $\times$8 \\
	\multirow{2}{*}{Wavelengths of instrumental artifacts (nm)} &  \\
      &1200 \\
      &1350 \\
      &1360$-$1450\\
      &1850 \\
      &1800$-$1950 \\
      &2000 \\
      Vendor-reported linearity range (dB) & 40 \\
      Directly measured linearity range (dB) & $>24$ \\
    \hline
    \end{tabular}
\end{center}
\end{table}

We also looked for the indication of large gaps detectable as ``IR bubbles'' \cite{1992JEMat..21..669M} with the coarse IR imaging technique.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic illumination source.  The lens was spaced from the detector plane by about XX mm, with the illumination source at a distance XXX from the lens.  The illumination source was gently diverging.  The result was a vignetted beam.  The target was located a distance XXX from the lens to provide imaging of the Si interface, with a field of view of roughly X $\times$ X degrees, or X $\times$ X mm.  This field of view sub-sampled the 4 inch diameter wafer, so we dithered the sample, collecting images.

%\begin{figure}[htbp]
%\centerline{\includegraphics[width=.8\columnwidth]{IMG_7662.JPG}}
%\caption{Photo of part VG15 with locations of defects marked.}
%\end{figure}

\section{Results}

\appendix
\section{Incoherent Multiple Reflections Transfer Matrix Method}
\label{sec:Append-IMRTMM}

@@

>>>>

The University of Texas diffractive optics group produces diffraction gratings on silicon using microlithographic techniques \cite{1998SPIE.3354..201J,2010SPIE.7739E.146W}.  These gratings are used either in transmission (i.e. as grisms \cite{2008SPIE.7014E..77D,2010SPIE.7739E.123G}), or in reflection (i.e. as immersion gratings, \cite{2007ApOpt..46.3400M,2010SPIE.7739E.146W,2012SPIE.8450E..2SG}.  These gratings operate throughout the infrared, from 1.2 to 38 $\mu$m, and find application in ground-based (IGRINS \cite{2010SPIE.7735E..54Y,2012SPIE.8450E..2SG}), airborne (SOFIA FORCAST \cite{2008SPIE.7014E..77D}), and space (JWST NIRCAM \cite{2005SPIE.5904...21B,2010SPIE.7739E.123G}) applications.  To date, our group has produced the gratings on large (up to 100 mm diameter and 30 mm thick) monolithic pieces of monocrystalline silicon and we shape the substrates by sawing, grinding, and polishing to form the final devices \cite{2010SPIE.7739E.146W}.  The next generation of astronomy and earth science spectrographs, however, will need optical devices larger and more precise than what can be made with existing technology.  The proposed Giant Magellan Telescope \cite{2012SPIE.8444E..1HJ} Near Infrared Spectrograph \cite{2006SPIE.6269E.143J,2010SPIE.7735E..87L} (GMTNIRS) will need immersion gratings with thicknesses up to 60 mm and lengths up to 200 mm.  Substrates of the required size cannot fit into conventional processing equipment like reactive ion etchers and electron-beam lithography tools.  Large, thick Si substrates are also prohibitively expensive to experiment on.  Silicon direct bonding \cite{1986JAP....60.2987S,2012SPIE.8450E..2TV} offers advantages by reducing the thickness and mass needed for the development of a thin (0.5$-$10 mm) Silicon diffractive optic that can later be bonded to a fine polished Si prism substrate.  Other groups producing SI diffractive optics are also experimenting with this approach (see \cite{2012SPIE.8450E..2TV} and \url{www.lumella.com}).

While it is possible that proper precautions to eliminate all micron-sized particles from both SI surfaces prior to bonding, the power-law nature of the particle size distribution in clean-room environments (\cite{Cooper_1986}) means that it is significantly more difficult to assure that no particles of the 10$-$60 nm size scale are present.  Small particles could be a concern if they act as "tent poles", ti create gaps where the bonding is incomplete.

There are two major concerns with Si direct bonding: the performance and robustness.  Specifically, what happens if the bonding process is imperfect and there are small gaps between the Si parts?  Does a gap degrade the performance of a final device, either through optical aberrations or transmission loss?  How mechanically robust is the final unit?  What is the strength of the bond?  Can the interface be broken by shearing, torquing, pulling, or thermal stress?  In practice, the figure of merit for robustness is whether the part can survive a shake test simulating the forcing experienced during a rocket launch or transport, or during cryogenic cycling, if the part will be used at low temperature.  Optically, the relevant parameters are the loss caused by the gap and the effective distortion of the transmitted wavefront.

We describe a technique for directly measuring gaps at silicon-silicon interfaces.  Other techniques for measuring and mapping gaps have already been described elsewhere \cite{1992JEMat..21..669M}, including IR interference, X-ray topography, and ultrasound microscopy.  The standard infrared interference technique is to examine the bond in reflection for the presence of Newton rings.  The presence and number of such rings can yield information about both the lateral and axial extend of any gap, as long as the gap along the direction of propagation is of the order of the vacuum wavelength or larger.  X-ray topography \cite{1992JEMat..21..669M,1994JaJAP..33....6H} exploits the change in optical density along lines of sight with gaps to allow precise and high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but provides only coarse estimates of the axial extent.  Ultrasound microscopy exploits the density change at the gap interface to produce reflections of sound waves and map the lateral gap dimensions but produces limited information about the axial dimension \cite{2000RScI...71.1869G}.

Here we describe an exploitation of the IR interference technique with new precision instrumentation in the near infrared wavelength range to measure the axial extent of gaps at the few nanometer size scale.  In this paper we show that gaps as small as 35 nm cause an unacceptable degradation in transmission performance.  Because the size scales probed are so small, they are comparable to the range over which molecular (van der Waals) forces are acting.  It is conceivable that the strength of the bond energy can be constrained from the \emph{in situ} measurement of the separation and that therefore the optical technique can provide not only a measure of the transmission but also an indication of the mechanical robustness of the bond.  By measuring both the spatial and axial extent of any gaps, the new metrology technique informs revisions of the bonding strategy that can improve both the optical and mechanical performance.

Our new technique for measurement of the axial extent of gaps exploits wavelength-dependence of reflection losses at the gap and the availability of commercial scanning monochromator systems with both high precision and good amplitude stability.  We first describe the theory behind the technique and predicted outcomes for various gap dimensions.  We then evaluate the technique experimentally on a variety of bonded samples.

\section{History of Si direct bonding}
\ref{secHistory}

@@

The majority of work on bonding was carried out from 1990-2000.  The key advances included studies that probed the surface energy as a function of anneal temperature, and various probes of the bond chemistry through multiply reflection infrared spectroscopy and mass spectroscopy.  A fairly complete story was then constructed which tied together all the disparate measurements.  No where in the literature did we find a technique similar to the one presented here.  See the table below for a list of papers on Si wafer bonding and a note on their contribution.

%\caption{Literature review}

@@

\begin{center}
\begin{longtable}{ |c | p{8cm} | p{8cm} |}

@@

\end{longtable}
\end{center}

@@

Our measurement technique exploits the large Fresnel reflection \cite{2001opt4.book.....H} at the Si-Air interface.  The refractive index of silicon at room temperature ranges from 3.55$-$3.45 from $\lambda = $ 1150$-$2500 nm \cite{2006SPIE.6273E..77F}, one of the largest refractive indices of any conventional material.  The Fresnel reflection is accordingly about 30\% per surface.  Popular optical materials with refractive indices in the range 1.5$-$1.7 with Fresnel reflection 4$-$7\% per surface will suffer less than Si from the effects described here.

Consider the following two conceivable pitfalls in Si direct bonding.  First, particles at the tens of nanometer size scale are trapped between the Si pucks during bonding.  This potential problem is especially acute for silicon direct bonding compared to bonding of popular low-index optical materials.  A typical particle surface flux is shown in Figure 5 of Cooper 1986\cite{Cooper_1986}.  The flux of particles is a step function of particle size.  For every micron sized particle causing Newton's rings, there will be as many as 100,000 more ten-nanometer sized particles.

The second pitfall to consider is that the stiff non-flat Si surfaces will not conform upon contact leaving the complement of the aberrations as an air gap. Similarly, surface roughness will also limit the bonding strength  \cite{JJAP.37.4197,2001JOptA...3...85G}.  The details of this aspect of the bonding process are outside the scope of this paper.

@@

\begin{table}[h!]
%\caption{UTexas Si bonding experiments \label{tab:bondexper}}
\begin{center}
    \begin{tabular}{ c c c c c c c}
    \hline
    Name & Bonded/DSP & Diameter & Thickness & Anneal Temp & Date & Pattern notes \\ 
    - & - & mm & mm & $^\circ$C & MM/DD/YYY & - \\ 
        \hline
    VG01 & Bonded  & 100  &$1$ $\times$ 2 & 23  & 2/13/2013 &  \\
    VG02 & Bonded & 100 & $3.3$ $\times$ 2 &  23  & 2/13/2013 &  \\
    VG03 & Bonded & 100 & $3.3$ $\times$ 2 &  23 & 2/14/2013 & 4 $\mu$m hole \\    
    VG04 & Bonded & 100 & $1$ $\times$ 2 &  23 & 2/14/2013 & 15 nm gap \\        
    VG05 & DSP & 100 & 1&  23 & 2/15/2013 & cleaved  \\
    VG06 & DSP & 100 & 3.5 &  23 & 2/15/2013 & \\
    VG07 & - & - & - &  - & - & \\
    VG08 & - & - & - &  - & - & \\
    VG09 & Bonded & 75 & 1.5? & 23 & 6/21/2013 & mesh C, 49 $\pm$6 nm \\
    VG15 & Bonded & 100 & 1.5? &  23 & 6/21/2013 & mesh C, M, F, 14 $\pm$2 nm\\
    VG17 & Bonded & 75 & 6.0? &  23 & 6/21/2013 & mesh C, 64 $\pm$6 nm\\
    VS20 & Bonded & 100 & 1.5? &  23 & 6/21/2013 & mesh C, M, F, 95 $\pm$5 nm\\
    \hline
    \end{tabular}
\end{center}
\end{table}

\section{Transmission measurement theory}
The Si-air\footnote{We imprecisely refer to the gap as an air gap, though in principle the gap has an unknown mixture of gases for example water vapor, or other vapors that were outgassed from reactions catalyzed by hydrocarbons.} interfaces of the in the bond gap make a Fabry-P\`erot etalon\cite{2007fuph.book.....S}.  The transmission through a Fabry-Perot depends on the wavelength of light, the reflectivity of the etalon sidewalls, and the size of the gap.  We assume transmission is at normal incidence, the index of the gap is 1.0, and the material is at room temperature.  We assume Si is nonabsorbing longward of $\lambda$ = 1150 nm.  We compute the reflectivity for an Si-air interface, which is a modest function of wavelength through the refractive index of Silicon, $n(\lambda, T)$ \cite{2006SPIE.6273E..77F}.  The computation is simply Fresnel's law at normal incidence:

$$
R = \frac{(n-1)^2}{(n+1)^2}
$$

The reflectivity is about 30\% in the wavelength range 1100 to 5000 nm.  The modest wavelength dependence is shown in the bottom curve of the figure below.  The coefficient of Finesse\cite{2007fuph.book.....S} encapsulates the Fabry-P\`erot etalon's dependence on the reflectivity.  We compute a coefficient of Finesse in the range of about 2.7$-$2.9 depending on the wavelength.

\subsection{Consideration of multiple incoherent reflections}
Multiple incoherent reflections occur within the Si substrate.  Treatment for multiple coherent reflections have been described in detail in the optics literature \cite{2007fuph.book.....S}.  The wave transfer matrix technique treats each dielectric interface as a matrix with elements relating the pre- and post- interface complex amplitudes in the left and right directions.  I adapted the wave transfer technique for incoherent interactions \cite{2002ApOpt..41.3978K}.  Specifically I constructed an incoherent wave transfer matrix whose elements relate the intensities (and not complex amplitudes) before and after an interface.  The details are worked out in Appendix \ref{sec:Append-IMRTMM}.  We computed the transmission for two scenarios- first a double side polished (DSP) Si puck with no air gap, and second a pair of bonded Si pucks with a gap thickness $d$.  The DSP Si puck with no air gap has a transmission equal to:

$$
T_{DSP} = \frac{2n}{1+n^2}
$$

which has an average value of about 53\%, as shown in the top curve of the Figure above.  Note that this transmission is above a na\"ive value of $T_{DSP}=(1-R^2)$, which does not take into account multiple incoherent reflections, and is therefore an underestimate.  If the Si wafer path length is less than the coherence length for the given spectral bandwidth, then the multiple reflections would interfere coherently, and the Si wafer would behave as a Fabry-P\`erot etalon.  We work out this behavior in the Appendix section X and show that in the limit $l>> \lambda^2 / \delta \lambda $ the spectral transmission tends to equation X.

For the case of a pair of bonded Si pucks with a gap thickness $d$, we normalize the transmission with a gap to the transmission with no gap to isolate the effect of the gap.  We call this normalized transmission the etalon transmission $T_{e}$.  It is derived in Equation \ref{eqn:Tetalon} in the Appendix:

$$T_{e} = \frac{n^2+1}{2 n F \sin ^2(2\pi \frac{d}{\lambda})+n^2+1} \nonumber $$

@@

\section{Substrate preparation}
It is important to clean the wafers to minimize the particle density, since particles cause gaps.  The substrates were silicon wafers.  The surface roughness was typically about 2 nm, as measured with a Wyko 2D noncontact microscope.  The substrate thicknesses ranged from 0.5 to 3.3 mm per substrate.  We classify the 3.3 mm thick substrates as ``pucks'' and the $\sim$1.0 mm thick substrates as ``wafers''.  The distinction between the two classes of substrate thickness is based on the anticipation that thin substrates will conform more easily to their bonding partner substrates.  There is also evidence that bond front propagation is much slower in thick pucks \cite{2007ApOpt..46.6793H} than in wafers.  We measured the large scale surface flatness of the pucks.  The Fisba2 interferometer has a 2 inch diameter beam at $\lambda=$632.8 nm.  We found a typical peak to valley surface flatness of 2 waves over the central 2 inch diameter.

\section{Measurement strategy, conditions, and details}
The measurements were performed with an Agilent Cary 5000 UV-Vis-NIR spectrophotometer \nocite{cary5000web}.  The typical measurement parameters and tool performance are listed below.  More details about the tool are available at the vendor's website.  Briefly, the tool has a double monochromator.  The double monochromator feature reduces scattered light, with the vendor reported linearity exceeding 40 dB.

The Cary 5000 has four operation modes: single front, single back, double, and double reverse.  We experimented with single front and double modes.  Double mode takes a reference spectrum simultaneously to the target spectrum.  In both single and double beam mode a baseline scan is taken of air.  The baseline spectrum is the raw spectral response of the system, which is automatically divided out of subsequent spectra.  In single beam mode the baseline measurements must be repeated about every half hour.  We estimated the uncertainty and measurement repeatability of single and double modes in the following way.  We computed the RMS error and measurement repeatability in a single measurement by computing the standard deviation of the beam with no sample in a series of spectra taken immediately after a 100\% baseline scan.  The predicted mean level should naturally be 100\%.  We found that the mean value in single mode drifted by about 0.2\%.  The standard deviation of the featureless spectrum was typically 0.02\%, with some portions of the spectrum demonstrating increased uncertainty, perhaps attributable to atmospheric absorption. In double beam mode the spectrograph performance was much better.  We put a double side polished Si wafer in the reference beam, and no sample in the target beam during the baseline scan.  In this way the transmission spectrum of a target Si wafer is automatically normalized.  The double beam mode effectively divides out variations in the mean transmission levels attributable to lamp drift at 0.2\% in single beam measurements.  The drift in mean transmission level is merely 0.02\% in double beam mode.  These values are accurate for measurements over a half hour timescale or less.  Over periods of hours the drift increases for both single and double beam mode, so a repeat baseline scan is sometimes necessary.  The need for repeat baselines is more acute for single mode.

@@

Table \ref{tab:tablenum3} shows some properties of the Cary5000.

%\caption{Summary of Cary 5000 Measurement parameters \label{tab:tablenum3}}
\begin{center}
    \begin{tabular}{l l}
        Parameter (Units) & Value \\ 
        Spectral sampling interval (nm) & 2.0 \\
        Spectral resolution (nm) & 5 \\
        Time per sample (s) & 0.1 \\
        Available measurement range (nm) & 175-3300 \\
        Typical measurement range (nm) & 1000-2500 \\
	Single mode RMS  (percent) & 0.02 \\
	Double mode RMS  (percent) & 0.002 \\
	Single mode mean drift (percent) & 0.20 \\
	Double mode mean drift (percent) & 0.02 \\
	Mean 0\% baseline correction (percent) & 0.03 \\
 	Beam size (mm x mm) & 2 x 8 \\
	Wavelengths of instrumental artifacts (nm) & stuff \\
        num1 & 1200 \\
        num2 & 1350 \\
        num3 & 1360-1450\\
        num4 & 1850 \\
        num5 & 1800-1950 \\
        num6 & 2000 \\
        Vendor-reported linearity range (dB) & 40 \\
        Directly measured linearity range (dB) & 24 \\
      \end{tabular}
\end{center}

We also looked for the indication of large gaps detectable as ``IR bubbles'' \cite{1992JEMat..21..669M} with the coarse IR imaging technique.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic white light illumination source.  We estimate that the lamp-detector combination was sensitive to wavelengths in the range 1.1-1.6 $\mu$m.  The field of view sub-sampled the 4 inch diameter wafers, so we dithered the sample.  No effort was made to flat field the detector, so copious detector non-uniformities and vignetting are present in our images.  It was easy to detect IR bubbles despite the artifacts since the bubbles translate with the wafer during dithering.  We detected IR bubbles in all 3 of the bonded Si pucks on which we recorded IR images.  The bubble areal density varied from about 4 per wafer (100 mm diameter) to over 20 per wafer.  Naturally, the sample of control solitary DSP (unbonded) wafers show no such IR bubbles, which are entirely attributable to the gaps between bonded substrtes as discussed in the literature (see Section \label{secHistory}).  Figure \ref{IRbubble} shows an example of one of the IR bubbles in which several bright and dark fringes are clearly discernible.

\begin{figure}[htbp]
\centerline{\includegraphics[width=.8\columnwidth]{figs/IR_bubble.pdf}}
\caption{Photo of part VG09 with transmission image of a bubble highlighted\label{IRbubble}.  The right side photo shows the 75 mm diameter wafer with blue Sharpie rectangles indicating the position of different spectral measurements (as detailed in the next section).  The yellow circle in the photo highlights the outline of the location of an IR bubble we detected.  The bubble shows a couple of fringes over the effective wavelength range 1.1-1.6 $\mu$m.  This fringe pattern is consistent with a gap comparable to this measurement wavelength.  The pointlike measurement artifacts, Sharpie marker, and vignetting are perceptible in the raw IR image.  Our interpretation of the IR bubbles is robust against these artifacts.  The area surrounding the IR bubble appears to be mostly devoid of large (wavelength-scale) gaps, since the transmission is fairly uniform, and consistent with the brightest fringe of the IR bubble.  }
\end{figure}

\section{Results of infrared spectroscopy of bonded Si pucks}

\appendix

\section{Appendix: X-ray Computed Tomography}
We acquired x-ray computed tomography of sample VG03.  The measurement details are listed in Table\ref{tab:tablenum4}.

%\caption{Summary of X-ray CT measurements for VG03 \label{tab:tablenum4}}
\begin{center}
    \begin{tabular}{ll}
    \emph{Parameter (Units)} & \emph{Value} \\ 
    Instrument& Xradia MicroCT \\
     Objective & 0.7X \\
    Accelerating voltage (kV) & 100 \\
        Power (W) & 10 \\
        Acquisition time (s) & 3.5 \\
        Number of views & 2521 \\
        Total slices & 469 \\
        Field of view diameter (pixels) & 988 \\
        Voxel lateral pixel resolution ($\mu$m) & 38.02 \\
        Estimate field of view (mm) & 37.5 \\
        Voxel axial pixel resolution ($\mu$m) & 0.5 \\
        Scan date & 2013 August 8 \\
        Sample thickness (mm) & 6.6 \\
     \end{tabular}
\end{center}

The axial resolution of the X-ray CT system is about 500 nm, which is much larger than the small ($\sim$ 15$-$90 nm) patterned gaps on many of our bonded samples.  Part VG03 has a cylindrical patterned gap with a $\sim$20 mm diameter and 4000 nm height.  We expected to see the cylindrical gap in about 8 resolution elements, given that the pattern was centered in the 25 mm diameter field of view.  We saw a circular feature that extended over 297 pixels, which corresponds to 11.3 mm, assuming the voxel lateral dimension from the table above, which is reported in the X-ray CT lab measurement details.  The feature...

\section{Incoherent Multiple Reflections Transfer Matrix Method}
\label{sec:Append-IMRTMM}

@@

\section{Appendix: High spectral resolution coherent interference of thin substrates}

Throughout this article we have assumed that $L >> \lambda^2 / \delta \lambda$, so that the interference of the substrate is safely incoherent.  Another way of stating this criterion is that the spectral bandwidth is large enough that high spectral frequency fringes average out in a composite spectrum of low spectral resolution.  If we relax this assumption so that the Si substrate thickness is comparable to or smaller than the coherence length, we would begin to observe high frequency fringes in the transmission spectra.  In this appendix section we work out the transmission spectrum for two scenarios.  First, we work out the case of a DSP Si wafer of thickness $L$, with no gap.  Second we work out the case of two bonded Si wafers of thicknesses $L_1$ and $L_2$.  For both cases we consider infinite spectral resolution, then convolve the spectra with a finite spectral resolution to show that the spectrum tends to that derived using an incoherent assumption.

\subsection{DSP Si wafer at high spectral resolution}
The Fabry-P\`erot etalon transmission is given in Equation \# X (currently, 5).  What we want to evaluate is:

$$
 <T_e> = \lim_{\sin^2(\delta/2) \to 0.5} \frac{1}{1+F\sin^2(\delta/2)} \\
 <T_e> = \frac{1}{1+F/2}
$$
